---
title: "Exploratory Analysis of Maternal Health Risk Factors in Bangladesh"
date: 2026-07-31
author: "Brototi Chakrabarty"
---


## Introduction

Maternal health risk assessment is important for identifying pregnant women who may require additional medical care. In this project, I  analyzed a maternal health risk dataset collected from hospitals and community clinics in rural Bangladesh. The dataset includes maternal age, systolic and diastolic blood pressure, blood sugar, body temperature, heart rate, and maternal risk level.

I  used R to compare low-, mid-, and high-risk groups, examine relationships among the clinical measurements, and determine whether principal component analysis revealed patterns among the maternal risk categories.

## Packages Used

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(here)

```


## Get the Data

```{r}
maternal_data <- read_csv(
  here("data", "Maternal Health Risk Data Set.csv"),
  show_col_types = FALSE
)
```
## Check the Data

```{r}
glimpse(maternal_data)
```
## Prepare the Data

```{r}
maternal_data <- maternal_data |>
  mutate(
    RiskLevel = str_to_lower(str_trim(RiskLevel)),
    RiskLevel = factor(
      RiskLevel,
      levels = c("low risk", "mid risk", "high risk"),
      ordered = TRUE))
```

```{r}
levels(maternal_data$RiskLevel)
```
## Questions

### 1. How many observations are classified as low, mid, or high risk?

To describe the distribution of maternal risk, I counted the number of observations in each risk category.

```{r}
maternal_data |>
  count(RiskLevel, sort = TRUE) |>
  ggplot(aes(
    x = RiskLevel,
    y = n,
    fill = RiskLevel
  )) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = n),
    vjust = -0.4
  ) +
  labs(
    title = "Number of Observations in Each Maternal Risk Category",
    x = "Maternal Risk Level",
    y = "Number of Observations"
  ) +
  theme_minimal()
```
Low risk was the most common category, with 406 observations (40.0%), followed by mid risk with 336 observations (33.1%) and high risk with 272 observations (26.8%). Thus, more than one-quarter of the dataset was classified as high risk.

### 2. How does maternal age differ across risk groups?

Next, I examined how maternal age varied across the low-, mid-, and high-risk categories.

```{r}
maternal_data |>
  ggplot(aes(
    x = RiskLevel,
    y = Age,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Maternal Age Across Risk Categories",
    x = "Maternal Risk Level",
    y = "Age"
  ) +
  theme_minimal()
```
The high-risk group had the highest median maternal age, while the low-risk group had the lowest. Although the distributions overlapped, older observations were more common in the high-risk category.

### 3. How do systolic and diastolic blood pressure differ across risk groups?

I then compared systolic and diastolic blood pressure across the three maternal risk categories.

```{r}
maternal_data |>
  select(RiskLevel, SystolicBP, DiastolicBP) |>
  pivot_longer(
    cols = c(SystolicBP, DiastolicBP),
    names_to = "BloodPressureType",
    values_to = "BloodPressure"
  ) |>
  ggplot(aes(
    x = RiskLevel,
    y = BloodPressure,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(
    ~BloodPressureType,
    scales = "free_y"
  ) +
  labs(
    title = "Blood Pressure Across Maternal Risk Categories",
    x = "Maternal Risk Level",
    y = "Blood Pressure"
  ) +
  theme_minimal()
```
Both systolic and diastolic blood pressure were higher in the high-risk group than in the low- and mid-risk groups. The separation was especially noticeable for diastolic blood pressure.

### 4. How does blood sugar differ across maternal risk groups?

Blood sugar distributions were evaluated across the low-, mid-, and high-risk groups.

```{r}
maternal_data |>
  ggplot(aes(
    x = RiskLevel,
    y = BS,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Blood Sugar Across Maternal Risk Categories",
    x = "Maternal Risk Level",
    y = "Blood Sugar"
  ) +
  theme_minimal()
```
Blood sugar was clearly higher in the high-risk group, with a median of approximately 11 compared with about 7 in the low- and mid-risk groups. The high-risk group also showed greater variability.

### 5. How do body temperature and heart rate differ across risk groups?

To assess additional vital signs, I compared body temperature and heart rate among the maternal risk categories.

```{r}
maternal_data |>
  select(RiskLevel, BodyTemp, HeartRate) |>
  pivot_longer(
    cols = c(BodyTemp, HeartRate),
    names_to = "Measurement",
    values_to = "Value"
  ) |>
  ggplot(aes(
    x = RiskLevel,
    y = Value,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(
    ~Measurement,
    scales = "free_y"
  ) +
  labs(
    title = "Body Temperature and Heart Rate Across Risk Categories",
    x = "Maternal Risk Level",
    y = "Observed Value"
  ) +
  theme_minimal()
```
Body temperature was generally similar across the three groups, although the mid- and high-risk groups showed a wider range. Heart rate was slightly higher in the high-risk group, and one unusually low value appeared in the low-risk group.

### 6. Which clinical measurements are most strongly associated with maternal risk level?

To identify the measurements most strongly associated with maternal risk, I converted the ordered risk categories to numeric scores: low risk as 1, mid risk as 2, and high risk as 3. I then calculated Spearman correlations.

```{r}
risk_associations <- maternal_data |>
  mutate(
    RiskScore = as.numeric(RiskLevel)
  ) |>
  summarise(
    across(
      c(
        Age,
        SystolicBP,
        DiastolicBP,
        BS,
        BodyTemp,
        HeartRate
      ),
      ~cor(
        .x,
        RiskScore,
        method = "spearman",
        use = "complete.obs"
      )
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Measurement",
    values_to = "Correlation"
  ) |>
  arrange(desc(Correlation))

risk_associations
```

```{r}
risk_associations |>
  ggplot(aes(
    x = reorder(Measurement, Correlation),
    y = Correlation
  )) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Clinical Measurements Associated With Maternal Risk Level",
    x = "Clinical Measurement",
    y = "Spearman Correlation"
  ) +
  theme_minimal()
```
  Blood sugar had the strongest positive correlation with increasing maternal risk level, followed by systolic blood pressure, diastolic blood pressure, and age. Heart rate and body temperature showed much weaker relationships.


### 7. Can PCA visually separate observations according to maternal risk category?

Finally, I performed principal component analysis using the six numeric clinical measurements. The variables were centered and standardized because they were measured on different scales.

```{r}
# Create the PCA model
pca_model <- maternal_data |>
  select(
    Age,
    SystolicBP,
    DiastolicBP,
    BS,
    BodyTemp,
    HeartRate
  ) |>
  prcomp(
    center = TRUE,
    scale. = TRUE
  )

# Calculate variance explained
pca_variance <- pca_model$sdev^2 / sum(pca_model$sdev^2)

pc1_percent <- round(pca_variance[1] * 100, 1)
pc2_percent <- round(pca_variance[2] * 100, 1)

# Create PCA scores
pca_data <- as_tibble(pca_model$x) |>
  mutate(
    RiskLevel = maternal_data$RiskLevel
  )

# Create the PCA plot
pca_data |>
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = RiskLevel
  )) +
  geom_point(
    alpha = 0.6,
    size = 2
  ) +
  labs(
    title = "PCA of Maternal Health Measurements",
    subtitle = paste0(
      "PC1 and PC2 explain ",
      round(pc1_percent + pc2_percent, 1),
      "% of the total variation"
    ),
    x = paste0(
      "Principal Component 1 (",
      pc1_percent,
      "%)"
    ),
    y = paste0(
      "Principal Component 2 (",
      pc2_percent,
      "%)"
    ),
    color = "Risk Level"
  ) +
  theme_minimal()
```

The PCA plot shows partial separation of the maternal risk categories. High-risk observations were more frequently located on the positive side of Principal Component 1, while many low- and mid-risk observations were concentrated near the center or negative side. However, substantial overlap remained among all three groups, indicating that the six clinical measurements did not completely separate patients by maternal risk category. One low-risk observation appeared far from the other points along Principal Component 2 and may represent an unusual record or outlier.


## Discussion
High-risk observations generally had higher age, blood pressure, and blood sugar. Blood sugar showed the strongest association with risk level, while body temperature and heart rate showed weaker differences. PCA showed partial separation among risk groups, but considerable overlap remained.

## Conclusions

Maternal risk level was most strongly associated with blood sugar and blood pressure. These results show associations only and do not prove causation.

## References

Ahmed, M. (2020). *Maternal Health Risk* [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C5DP5D

---

**Last updated:** July 2026
