---
title: "Exploratory Analysis of Maternal Health Risk Factors in Bangladesh"
date: 2026-06-04
author: "Brototi Chakrabarty"
---

# Final Report: Exploratory Analysis of Maternal Health Risk Factors in Bangladesh

## Introduction

Maternal health risk assessment is important for identifying pregnant women who may require additional medical care. In this project, I will analyze a maternal health risk dataset collected from hospitals and community clinics in rural Bangladesh. The dataset includes maternal age, systolic and diastolic blood pressure, blood sugar, body temperature, heart rate, and maternal risk level.

I will use R to compare low-, mid-, and high-risk groups, examine relationships among the clinical measurements, and determine whether principal component analysis can reveal patterns among the maternal risk categories.

## Packages Used

```{r}
library(tidyverse)
library(here)
```
## Methods

The dataset will be imported and analyzed using R and the tidyverse package. I will examine the structure and quality of the data, summarize the clinical measurements, and create bar charts, boxplots, scatterplots, and correlation plots. Principal component analysis will also be used to explore whether the maternal risk categories show distinct patterns.

## Get the Data

```{r}
maternal_data <- read_csv(
  here("data", "Maternal Health Risk Data Set.csv"),
  show_col_types = FALSE
)
```
## Check the Data

```{r}
glimpse(maternal_data)
```
## Prepare the Data

```{r}
maternal_data <- maternal_data |>
  mutate(
    RiskLevel = str_to_lower(str_trim(RiskLevel)),
    RiskLevel = factor(
      RiskLevel,
      levels = c("low risk", "mid risk", "high risk"),
      ordered = TRUE))
```

```{r}
levels(maternal_data$RiskLevel)
```
## Questions

### 1. How many observations are classified as low, mid, or high risk?

First, I will count the number of observations in each maternal risk category.

```{r}
maternal_data |>
  count(RiskLevel, sort = TRUE) |>
  ggplot(aes(
    x = RiskLevel,
    y = n,
    fill = RiskLevel
  )) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(
    aes(label = n),
    vjust = -0.4
  ) +
  labs(
    title = "Number of Observations in Each Maternal Risk Category",
    x = "Maternal Risk Level",
    y = "Number of Observations"
  ) +
  theme_minimal()
```
Low-risk observations were the most common, with 406 records (40.0%). Mid-risk observations included 336 records (33.1%), while high-risk observations included 272 records (26.8%). Although the high-risk group was the smallest category, it still represented more than one-quarter of the dataset.

### 2. How does maternal age differ across risk groups?

Maternal age will be compared across the low-, mid-, and high-risk categories.

```{r}
maternal_data |>
  ggplot(aes(
    x = RiskLevel,
    y = Age,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Maternal Age Across Risk Categories",
    x = "Maternal Risk Level",
    y = "Age"
  ) +
  theme_minimal()
```
Maternal age differed across the three risk categories. The high-risk group had a higher median age than the low- and mid-risk groups, although the age distributions overlapped. This suggests that higher maternal age is associated with increased risk classification in this dataset.

### 3. How do systolic and diastolic blood pressure differ across risk groups?

Systolic and diastolic blood pressure will be compared among the three maternal risk categories.

```{r}
maternal_data |>
  select(RiskLevel, SystolicBP, DiastolicBP) |>
  pivot_longer(
    cols = c(SystolicBP, DiastolicBP),
    names_to = "BloodPressureType",
    values_to = "BloodPressure"
  ) |>
  ggplot(aes(
    x = RiskLevel,
    y = BloodPressure,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(
    ~BloodPressureType,
    scales = "free_y"
  ) +
  labs(
    title = "Blood Pressure Across Maternal Risk Categories",
    x = "Maternal Risk Level",
    y = "Blood Pressure"
  ) +
  theme_minimal()
```
Both systolic and diastolic blood pressure were higher in the high-risk group than in the low- and mid-risk groups. The median systolic blood pressure was approximately 130 mmHg in the high-risk group compared with about 120 mmHg in the other groups, while the median diastolic blood pressure was approximately 90 mmHg compared with about 75 mmHg. Although the distributions overlapped, the results suggest that higher blood pressure is associated with a higher maternal risk classification in this dataset.

### 4. How does blood sugar differ across maternal risk groups?

Blood sugar measurements will be compared among the low-, mid-, and high-risk groups.

```{r}
maternal_data |>
  ggplot(aes(
    x = RiskLevel,
    y = BS,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Blood Sugar Across Maternal Risk Categories",
    x = "Maternal Risk Level",
    y = "Blood Sugar"
  ) +
  theme_minimal()
```
Blood sugar levels were clearly higher in the high-risk group than in the low- and mid-risk groups. The high-risk group had a median blood sugar value of approximately 11, while the medians for the low- and mid-risk groups were around 7. Although some high blood sugar outliers were present in the mid-risk group, the overall pattern suggests that higher blood sugar is associated with a higher maternal risk classification in this dataset.

### 5. How do body temperature and heart rate differ across risk groups?

Body temperature and heart rate will be compared among the maternal risk categories.

```{r}
maternal_data |>
  select(RiskLevel, BodyTemp, HeartRate) |>
  pivot_longer(
    cols = c(BodyTemp, HeartRate),
    names_to = "Measurement",
    values_to = "Value"
  ) |>
  ggplot(aes(
    x = RiskLevel,
    y = Value,
    fill = RiskLevel
  )) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(
    ~Measurement,
    scales = "free_y"
  ) +
  labs(
    title = "Body Temperature and Heart Rate Across Risk Categories",
    x = "Maternal Risk Level",
    y = "Observed Value"
  ) +
  theme_minimal()
```
Body temperature was similar across the three risk groups, with most observations close to 98°F. The mid- and high-risk groups showed a wider range of temperatures than the low-risk group. Heart rate was slightly higher in the high-risk group, but the distributions overlapped considerably. One unusually low heart-rate value was present in the low-risk group and may represent a recording error. Overall, body temperature and heart rate showed weaker differences across risk categories than blood sugar and blood pressure.

### 6. Which clinical measurements are most strongly associated with maternal risk level?

The ordered risk categories will be converted to numeric scores, with low risk equal to 1, mid risk equal to 2, and high risk equal to 3. Spearman correlations will then be calculated.

```{r}
risk_associations <- maternal_data |>
  mutate(
    RiskScore = as.numeric(RiskLevel)
  ) |>
  summarise(
    across(
      c(
        Age,
        SystolicBP,
        DiastolicBP,
        BS,
        BodyTemp,
        HeartRate
      ),
      ~cor(
        .x,
        RiskScore,
        method = "spearman",
        use = "complete.obs"
      )
    )
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "Measurement",
    values_to = "Correlation"
  ) |>
  arrange(desc(Correlation))

risk_associations
```

```{r}
risk_associations |>
  ggplot(aes(
    x = reorder(Measurement, Correlation),
    y = Correlation
  )) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Clinical Measurements Associated With Maternal Risk Level",
    x = "Clinical Measurement",
    y = "Spearman Correlation"
  ) +
  theme_minimal()
```
Blood sugar showed the strongest positive association with increasing maternal risk level, followed by systolic blood pressure, diastolic blood pressure, and age. These results describe associations and do not demonstrate causation.

### 7. Can PCA visually separate observations according to maternal risk category?

Principal component analysis will be performed using the six numeric clinical measurements. The variables will be centered and standardized because they use different measurement scales.

```{r}
pca_model <- maternal_data |>
  select(
    Age,
    SystolicBP,
    DiastolicBP,
    BS,
    BodyTemp,
    HeartRate
  ) |>
  prcomp(
    center = TRUE,
    scale. = TRUE
  )
```

```{r}
pca_data <- as_tibble(pca_model$x) |>
  mutate(
    RiskLevel = maternal_data$RiskLevel)

pca_data |>
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = RiskLevel
  )) +
  geom_point(
    alpha = 0.6,
    size = 2
  ) +
  labs(
    title = "PCA of Maternal Health Measurements",
    subtitle = "Observations are colored by maternal risk category",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Risk Level"
  ) +
  theme_minimal()
```
The PCA plot shows partial separation of the maternal risk categories. High-risk observations were more frequently located on the positive side of Principal Component 1, while many low- and mid-risk observations were concentrated near the center or negative side. However, substantial overlap remained among all three groups, indicating that the six clinical measurements did not completely separate patients by maternal risk category. One low-risk observation appeared far from the other points along Principal Component 2 and may represent an unusual record or outlier.



## Discussion

What do your results mean? How do they relate to your research question? What limitations exist? What would you do next?

## Conclusions

Summarize your findings and their implications.

## References



---

**Last updated:** June 2026
